---
output: 
  pdf_document:
    keep_tex: true
bibliography: moose_parasites.bib
header-includes:
  - \usepackage{float}
  - \usepackage{subcaption}
---

\vspace*{\stretch{1.0}}
\begin{center}
  \Large{\textbf{Evaluation of infection risk of two endoparasites to moose (Alces alces) in the Adirondack Park of New York}}\\
  \vspace{5mm}
  \large{Alec Wong\textsuperscript{1}, Krysten Schuler\textsuperscript{1}, Angela Fuller\textsuperscript{1,2,3}}\\
  \vspace{5mm}
  \normalsize{\textsuperscript{1}Cornell University}\\
  \normalsize{\textsuperscript{2}New York Cooperative Fish and Wildlife Research Unit}\\
  \normalsize{\textsuperscript{3}U.S. Geological Survey}\\
  \vspace*{\fill}
\end{center}

\newpage

```{r, include=F}
library(ggplot2)
library(unmarked)
library(dplyr) ; `%<>%` = magrittr::`%<>%`
library(reshape2)
library(knitr)
library(kableExtra)
library(inlabru)

knitr::opts_chunk$set(echo = FALSE)

source('../DiseaseAnalysis/functions.R')

```

# Abstract

Endoparasites *Parelaphastrongylus tenuis* and *Fascioloides magna* were
attributed to 19% and 21% of mortalities in New York in 2015, respectively, and
may be contributing factors to limiting moose (*Alces alces*) populations in the
Adirondacks region. Field surveys were conducted to collect white-tailed deer
(*Odocoileus virginianus*) fecal pellets under distance sampling protocol both
to measure relative deer density and to diagnose the infection status of the
pellets by *P. tenuis* and *F. magna*. Over the three years of data collection,
41.7% of fecal pellets collected ($N=656$) contained *F. magna* eggs, and 34.7%
of fecal pellets contained L1 *P. tenuis* larvae. We attempted to quantify the
geospatial distribution of parasite infection using landscape variables such as
habitat, geomorphon structure, elevation, and spatial trend, but goodness-of-fit
tests indicated poor fit of the model structures we evaluated, whether assessed
as continuous intensity of infection (Poisson, negative binomial), or binary
infection status (binomial). We conclude that the data are not extensive enough
to quantify relationships between parasite infection in white-tailed deer fecal
pellets and geospatial effects.

# Introduction

Moose (*Alces alces*) populations are in decline in parts of their range in
North America such as in Minnesota, Vermont, and New Hampshire in the United
States of America and Nova Scotia, Alberta, Saskatchewan, Manitoba, and Ontario
in Canada [@Lankester2018; @Timmermann2017]. Some researchers point to parasite
infection from liver fluke (*Fascioloides magna*) and/or brainworm
(*Parelaphastrongylus tenuis*) as contributing factors to these declines
[@Murray2006; @Lankester2010; @Lankester2018; @Wunschmann2015], although this
claim is controversial [@Lenarz2009] and in some areas may coincide with other
factors such as depredation or low reproduction rates that may disguise effects
of parasite infection [@SHANNONBARBERMEYER2016; @Kuzyk2018]. Nevertheless, brain
worm and liver fluke cause morbidity and mortality in moose populations and
should be monitored.

Infections by *P. tenuis* are benign to white-tailed deer (*Odocoileus
virginianus*), but cause severe neurological disorders in moose caused by
migration of larvae through the central nervous system that often end in death
[@Pybus2001; @Lankester2010]. *Parelaphastrongylus tenuis* is acquired via
ingestion of gastropod intermediate hosts, in which *P. tenuis* larvae have molt
from L1 to L2, and L3 stages over approximately 4 weeks. Gastropods become
infected by ingesting the mucuosal layer surrounding white-tailed deer fecal
pellets, or from the soil over which the L1-stage larvae have been washed by
precipitation.

*Fascioloides magna* infection damage manifests as lesions in the hepatic tissue
due to migration of immature flukes or fibrous encapsulation of adult flukes
[@Pybus2001]. Again, moose are thought to become infected through ingestion of
gastropod intermediate hosts. A major distinction is that the gastropod hosts
are aquatic snails (typically within the genus *Lymnaea*) that become infected
by free-swimming miracidia introduced to the water as eggs in white-tailed deer
feces [@Pybus2001; @Peterson2013; @Malcicka2015]. The meracidia develop within
the snail, ultimately generating cercariae that encyst as metacercariae on
aquatic vegetation. Because moose select aquatic vegetation as a significant
proportion of their diet, *F. magna* metacercariae are presumably commonly
consumed by moose.

In the Adirondacks of New York, moose populations have increased since their
extirpation in the late 19th century, but their populations have not seen the
same growth during the 20th century as their New England counterparts
[@Timmermann2017]. Recent estimates of moose population size indicate that there
are approximately 600 to 1000 moose in the Adirondacks [@Wong2019], a range
which does not indicate substantial change from previous assertions of
population size [@Hickey2008]. Factors that may be limiting the population are
not established, but the focal parasites *P. tenuis* and *F. magna* have been
observed in moose necropsies with moderate frequency (REF Schuler and/or DEC).

The New York State Department of Environmental Conservation (DEC) has been
gathering information on moose health from incidental mortalities (n = 117)
since 2000 and the Animal Health Diagnostic Center (AHDC) started live moose
testing in 2015. Moose in New York State appear to be substantially impacted by
several parasites excepting winter ticks, which are more common in New England
states. Over 19% of mortalities were attributed to P. tenuis, 21% of moose
carcasses examined were infected with liver flukes (Fascioloides magna).
Migrating parasites cannot be detected in live moose, but were likely a problem
as New York moose had 5-6x higher eosinophil (white blood cell) counts than New
Hampshire or Maine moose.

The geospatial distribution of infection risk to moose is a concern for
management, and we attempt to quantify this by first assessing the spatial
patterns in infection of white-tailed deer fecal matter.

# Materials and Methods

## Study area

We surveyed within the Adirondack Park of the state of New York. It is an area
of approximately 24,000 $\text{km}^2$, composing approximately 17% of the state
of New York. Elevation ranges from 100m - 1600m. The Adirondack Park contains
primarily upland forest habitat dominated by such species as American beech
(*Fagus grandifolia*), red spruce (*Picea rubens*), hemlock (*Tsuga
canadensis*), sugar maple (*Acer saccharum*), striped maple (*Acer
pensylvanicum*), and balsam fir (*Abies balsamea*). Approximately 20% of the
park is composed of wetland habitats including open river corridors, floating
bogs, and large open bogs dominated by conifers [@Hickey2008].

## Field data collection

White-tailed deer fecal pellet groups were collected opportunistically during
the years of 2016 and 2017 for the purpose of parasite quantification during a
larger survey investigating the population size of moose in the park. The moose
study surveyed triangular transects approximately 3 km long, selected by cluster
random sampling. The clustering design of the transects were optimized for a
spatial capture-recapture survey, including 3 transects per cluster separated by
approximately 2-4 km.


\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[height=500px]{img/sitePlot.png}
\caption{Transect locations within the Adirondack Park for the 2016 survey.}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[height=500px]{img/sitePlot2017.png}
\caption{Transect locations within the Adirondack Park for the 2017 survey.}
\end{subfigure}
\caption{\label{fig:sites}Transect locations within the Adirondack Park for the 2017 survey. Each dot represents the location of one transect, and each cluster of transects is bounded by a box. The inlaid plot indicates the extent of the Adirondack Park relative to the state of New York.}
\end{figure}


In 2018, a single-observer distance sampling protocol was implemented to
quantify deer fecal group abundance in addition to measuring parasite intensity.
Along each moose transects, 6 distance sampling transects each 100m long were
spaced 200m apart from each other. **INSERT FIGURE FOR DEER TRANSECTS** A total
of 150 transects were sampled in 2018. At the specified start point, field
technicians extended a 100m tape to serve as the center line, and made
observations along its length. Any number of scat pellets were considered to
constitute a fecal group, and the approximate number was recorded. When a fecal
group was detected, its distance perpendicular from the center line, and
parallel along the center line was recorded. The perpendicular distance is the
focal variable of distance sampling, but the parallel distance provided highly
precise location coordinates of the scats -- because the start point of the
transect was recorded with greater accuracy (due to longer GPS averaging time),
the coordinates of the scats could be refined by translating the start point in
space by the angle of the transect, and the parallel distance away from the
start point.

Fifty-seven white-tailed deer fecal samples were collected in 2016, 391 samples
were collected in 2017, and 174 samples were collected in 2018.

## Parasitological analyses

The fecal samples were analyzed by the Cornell Animal Health and Diagnostic
Center using several methods. For *F. magna*, Flukefinder, a modified Baermann
technique, and fecal quantitative flotation were used. For *P. tenuis*, the
modified Baermann, and fecal quantitative flotation methods were implemented.
After 2016, the performant method for *F. magna* was determined to be
Flukefinder, and for *P. tenuis* the performant method was the modified Baermann
technique. This was assessed by **HOW? ASK KRYSTEN**, and validated post-hoc to
have the greatest detection probability by an occupancy analysis.

### Flukefinder

*Needs to be updated by relevant members; Flukefinder methodology not in shared folder*

### Modified Baermann

### Fecal flotation

## Statistical Models

For analysis of relative risk of infection to moose, we implemented a
hierarchical Bayesian modeling framework analyzed with the `R` package `INLA`,
which uses integrated nested Laplace approximation to approximate the posterior
distribution of model parameters.

### Parasite intensity model

Let $\widehat{\lambda}(s)$ be the estimated per-fecal-group-sample abundance
(or, intensity in the epidemiological sense). Let $S = \{s_i, i = 1, ..., n\}
\subset \Omega$ be the $n$ distinct locations where white-tailed deer fecal
groups were observed within the observation region $\Omega \subseteq
\mathbb{R}^2$, and let $y_i$ be the observed parasite count within the fecal
sample.

Parasites observed within each scat sample are $\ge 0$, and due to the high
prevalence of 0's, we considered a negative binomial distribution for the
response in addition to the Poisson. On any bounded region $B \subseteq \Omega$,
observations are thus modeled as follows:

\begin{equation}
\begin{gathered}
y_i \sim \text{Poisson}(\mu_i) \\
\log(\mu_i) = \beta_0 + \beta_1*x_1(s_i) + ... + \beta_p*x_p(s_i) + \Psi_1(s_i) \\
\Psi_1(s_i) \sim GP(0, \Sigma)
\end{gathered}
\end{equation}

Where, $y_i$ represent the parasites observed per fecal group sample, $\mu_i$ is
the mean function modeled dependent upon spatial covariates $x_j(s)$, and
$\Psi_1(s)$ is a Gaussian process with mean 0 and covariance matrix
$\Sigma$. The covariance function used is the Matern; with $\nu = \alpha - d/2$,
and $\alpha = 2; d = 2$:

\[
\text{cov}[\Psi(s), \Psi(s')] = \sigma_{\Psi}^2 \  \kappa ||s' - s|| \ \mathcal{K}_1(\kappa ||s' - s||)
\]

where $\tau, \kappa$ are variance and range scaling parameters, respectively,
$\sigma_{\Psi}^2 = 1/(4\pi\kappa^2\tau^2)$ is the marginal variance, and
$\mathcal{K}_1$ is the modified Bessel function of the second kind and order 1
(See @Lindgren2015 for the specific representation and estimation of the spatial
random effects within INLA). We selected this error model to account for the
fact that the spatially clustered sampling is likely to obtain multiple samples
from the same individuals, inducing local spatial autocorrelation. While
empirical variograms do not demonstrate autocorrelation at any distance for *F.
magna* response values, there exists observed autocorrelation for *P. tenuis* at
the scale of approximately 1200 m.

Additionally, we assessed the fit of models treating the response as a binary
random variable modeled under the binomial distribution and zero-inflated
binomial distribution, with the same model of spatial dependence instead using
the appropriate link function.

#### Covariates on parasite intensity

In our model, we consider the following covariate effects on parasite intensity.
Coordinate data (UTM Easting and Northing) are included to model broad-scale
spatial trend. We considered a derived variable of precipitation representing
the total accumulation of precipitation in the year prior to the data collection
date, measured in inches. Precipitation affects the survival of *P. tenuis* and
*F. magna* outside of a host (i.e. as L1 larvae, and eggs in feces,
respectively), and so it is anticipated that deer would be more likely to be
infected in areas of higher rainfall [@Pybus2001]. Additionally, gastropod
activity is inhibited by drought, and so higher rainfall would suggest higher
risk of infection due to a greater availability of gastropods to be ingested in
the case of *P. tenuis* [@Lankester2018]. We used the spatially interpolated
PRISM dataset for precipitation [@PRISM]. We considered another derived variable
for snowfall accumulation in the snowfall season prior to the data collection
date (that is, from September until April prior to the date of survey). Similar
to precipitation, we anticipate that areas that experience low snowfall in the
year prior to the survey date will exhibit a higher parasite intensity due to
increased parasite survival, and increased activity of gastropod intermediate
hosts. The National Weather Service's National Snowfall Analysis dataset was
used to obtain spatially-referenced snowfall amounts.

Local terrain basins where water may collect were considered as a potentially
important effect particularly for *F. magna*, as the life cycle of the parasite
has a period during which metacercariae encyst on aquatic vegetation and once
matured infect aquatic snail intermediate hosts. We hypothesized that intensity
of infection for *F. magna* will be greater in basins than on slopes or peaks,
and we assessed this by the geomorphon tool in QGIS, assessed at a resolution of
40km (1600 squared-kilometers). This resolution was selected because it
maximized the variation in the geomorphon classification distribution relative
to our sample locations.

### INLA settings

#### Priors

We used default priors set by INLA for the covariate coefficients; that is:

$$
\beta_p \sim \text{Normal}(0, 0.001)
$$

where the second argument of the Normal distribution is specified on the
*precision* scale.

The priors for the Matern covariance parameters were specified using INLA's
penalized complexity prior construction for the spatial range parameter $\rho$
and marginal standard deviation $\sigma_0$. The joint density of the prior is:

\begin{equation}
\begin{gathered}
p(\rho, \sigma_0) = \frac{(d * R * S)}{2} \rho^{-1 - \frac{d}{2}} \exp(-R \rho^{\frac{-d}{2}}) \exp(-S \sigma_0) \\
R = -\log(p_\rho)\rho_0^{\frac{d}{2}} \\
S = -\frac{\log(p_\sigma)}{\sigma_0}
\end{gathered}
\end{equation}

Critically, $p_\rho$ and $p_\sigma$ are defined such that:

$$
\begin{gathered}
p_\rho = P(\rho < \rho_0) \\
p_\sigma = P(\sigma > \sigma_0)
\end{gathered}
$$

In so doing, the prior specification can be defined on intuitive scales, setting
a lower and upper limit for $\rho$ and $\sigma_0$, respectively. We selected
$\rho_0 = 100$ (meters) and $\sigma_0 = 1$, and $p_\rho = p_\sigma = 0.01$
reasoning that spatial autocorrelation < 100m is indistinguishable from white
noise, and the marginal standard deviation would not be much larger than the
sample standard deviation.


#### Mesh

A mesh is required to fit Gaussian random effect models under the INLA
framework. The `meshbuilder` tool provided by package `INLA` was used to
visually select settings that fulfilled the following criteria (see Figure
\ref{fig:boundaries}):

* Relatively fine mesh resolution (< 1km) in the vicinity of sampled data
* Relatively low mesh resolution far from sampled data (>5km)


```{r loadmesh, message=F, include = F}
load('../DiseaseAnalysis/model_outputs/adk_mesh.Rdata')
adkbound = rgdal::readOGR(dsn = '../../GIS/adkParkPoly/adkParkPoly.shp')
load('../DiseaseAnalysis/data_no_na.Rdata')
```

```{r plotmesh, cache = T, fig.cap="\\label{fig:boundaries}Boundary line of the Adirondack Park (blue), and the mesh constructed for INLA (red).", fig.pos = "H", fig.height=3.5}
coords = data.frame(adkbound@polygons[[1]]@Polygons[[1]]@coords)
colnames(coords) = c('x','y')

ggplot() + 
  ggmeshcust(adk_mesh) + 
  geom_path(data = coords, aes(x = x, y = y, color = 'Adirondack boundary'), size = 1.3) + 
  geom_point(data = data, aes(x = Easting_real, y = Northing_real, shape = 'datapoint')) + 
  scale_colour_manual(name = 'Boundaries', 
                      values = c('Adirondack boundary'='blue', 
                                 'interior'           ='red'),
                      labels = c('Adirondack Boundary', 'Interior (fine) mesh boundary')) + 
  scale_shape_manual(name = "Sampled data", values = 1, labels = "Sampled data") +
  guides(shape = guide_legend(title = "Points")) + 
  coord_equal() + 
  theme_bw() + theme(
    panel.grid = element_blank(),
    axis.title = element_blank()
  )

```


### Model Specifications

Six models were applied using the covariates developed above upon response
variables for both *F. magna* and *P. tenuis*.

\begin{table}[H]

\caption{\label{tab:models} Model specifications and their covariates}
\centering
\begin{tabular}[H]{ll}
\toprule
Model & Formula \\
\midrule
Null Model & . \\
Full Model & \makecell[l]{Easting + Northing + Precipitation + Snow + Geomorphon + Elevation + \\ Gaussian Process (GP)} \\
Fixed Effects Model & Easting + Northing + Precipitation + Snow + Geomorphon + Elevation \\
Fixed Effects Parasite Survival Model & Precipitation + Snow + Geomorphon \\
Two-scale spatial model & Easting + Northing + GP \\
Two-scale spatial model and elevation & Easting + Northing + Elevation + GP \\
\bottomrule
\end{tabular}
\end{table}


# Results

We observed 41.7% positive diagnoses of *F. magna* over the three years of data
collection ($N = 656$), with yearly positive results being 69% in 2016, ($N =
80$), 37.8\% in 2017 ($N = 402$), and 38.5\% in 2018 ($N = 174$). For *P.
tenuis* we observed 34.6% infection prevalence among the fecal samples tested
over all three years, with yearly positive results being 48.8% in 2016 ($N =
80$), 23.9\% in 2017 ($N = 402$), and 52.6\% in 2018 ($N = 174$) (Table \ref{tab:posTable}).

```{r posTable, message=FALSE}
metadata = read.csv(file = '../DiseaseAnalysis/metadata_adjusted.csv', stringsAsFactors = F)

ff = read.csv(file = '../DiseaseAnalysis/fecal_flukefinder.csv', stringsAsFactors = F)
mb = read.csv(file = '../DiseaseAnalysis/fecal_MB.csv', stringsAsFactors = F)
quant = read.csv(file = '../DiseaseAnalysis/fecal_quant.csv', stringsAsFactors = F)

m = metadata %>% left_join(ff %>% select(PK, Total_eggs)) %>% rename(fmagna_ff = Total_eggs) %>% 
  left_join(mb %>% select(PK, Total_larvae_dsl, total_eggs_fmagna)) %>% rename(dsl_mb = Total_larvae_dsl, fmagna_mb = total_eggs_fmagna) %>% 
  left_join(quant %>% select(PK, Fascioloides_magna, Protostrongylid_DSL)) %>% rename(fmagna_quant = Fascioloides_magna, dsl_quant = Protostrongylid_DSL)

m$fmagna_ff_y = m$fmagna_ff > 0
m$fmagna_mb_y = m$fmagna_mb > 0
m$fmagna_quant_y = m$fmagna_quant > 0

m$dsl_mb_y = m$dsl_mb > 0
m$dsl_quant_y = m$dsl_quant > 0

counts_ff = table(m$fmagna_ff_y, m$Year) %>% as.data.frame(stringsAsFactors = F)
counts_dsl = table(m$dsl_mb_y, m$Year) %>% as.data.frame(stringsAsFactors = F)

counts_ff$Species = 'F. magna'
counts_dsl$Species = 'P. tenuis'

p_ff = table(m$fmagna_ff_y, m$Year) %>% 
  prop.table(margin = 2) %>% 
  as.data.frame(stringsAsFactors = F) %>%
  rename(Proportion = Freq) %>% 
  mutate(Species = "F. magna")


p_dsl = table(m$dsl_mb_y, m$Year) %>% 
  prop.table(margin = 2) %>% 
  as.data.frame(stringsAsFactors = F) %>%
  rename(Proportion = Freq) %>% 
  mutate(Species = "P. tenuis")


table = rbind.data.frame(counts_ff, counts_dsl) %>% 
  rename(Diagnosis = Var1, Year = Var2, Count = Freq) %>% 
  mutate(Diagnosis = ifelse(Diagnosis, "Positive", "Negative")) %>% 
  mutate(Proportion = c(p_ff$Proportion, p_dsl$Proportion) %>% round(3))

c_table = table %>% reshape2::dcast(Diagnosis + Species ~ Year, value.var = c("Count")) %>% arrange(desc(Diagnosis))
p_table = table %>% reshape2::dcast(Diagnosis + Species ~ Year, value.var = "Proportion") %>% arrange(desc(Diagnosis))

j_table = cbind.data.frame(c_table, p_table[, c(3:5)])

knitr::kable(j_table, format = 'latex', booktabs = T, caption = "\\label{tab:posTable} Raw frequencies and proportions of diagnoses performed on white-tailed deer fecal samples between 2016 and 2018.") %>% 
  collapse_rows(columns = 1, latex_hline = 'major', valign = 'middle') %>%
  kable_styling(latex_options = c('HOLD_position')) %>% 
  add_header_above(c(" " = 2, "Counts" = 3, "Proportions" = 3))

```

## Model results

Model comparison was implemented by comparison of WAIC values produced by INLA.
This information criterion is a measure of *relative* out-of-sample predictive
accuracy, and lower WAIC values indicate greater relative performance.

Model results for the Poisson response distribution are not included in the
table because all AIC values are an order of magnitude larger than those of the
negative-binomial response distribution.

```{r aicvals, warning = F, include = F}
load('../DiseaseAnalysis/model_outputs/aicvals_fmagna.Rdata')
load('../DiseaseAnalysis/model_outputs/aicvals_ptenuis.Rdata')

model_matches = data.frame(fitnames = c("null_model", 
                                        "full_model", 
                                        "red_mod_linear_all_cov", 
                                        "red_mod_survival",
                                        "red_mod_spatial", 
                                        "red_mod_spatial_elev"),
                           pubnames = c("Null Model", 
                                        "Full model", 
                                        "Fixed effects model", 
                                        "Fixed Effects Parasite Survival Model", 
                                        "Two-scale spatial model", 
                                        "Two-scale spatial model and elevation"),
                           stringsAsFactors = F)

aic_vals_fmagna =
  aic_vals_fmagna %>% left_join(model_matches, by = c("model" = "fitnames"))

aic_vals_ptenuis =
  aic_vals_ptenuis %>% left_join(model_matches, by = c("model" = "fitnames"))

```

### *F. magna*

Among the models fit to the *F. magna* data, the best-supported model was the
global model, followed by the second-best model with a delta-WAIC value of 
`r (aic_vals_fmagna$waic %>% diff)[1] %>% round(2)`.


```{r fmagnawaic}

kable(x = aic_vals_fmagna %>% select(pubnames, waic), format = 'latex', booktabs = T, digits = 2, col.names = c("Model Name", "WAIC"))

```

### *P. tenuis*

Among the models fit to the *P. tenuis* data, the best-supported model was the
reduced 'two-scale spatial model', which had only easting, northing, and the 
Gaussian process error model. The next best performing model followed with a 
delta-WAIC value of `r (aic_vals_ptenuis$waic %>% diff)[1] %>% round(2)`.

```{r ptenuiswaic}

kable(x = aic_vals_ptenuis %>% select(pubnames, waic), format = 'latex', booktabs = T, digits = 2, col.names = c("Model Name", "WAIC"))

```

### Fixed effects parameter estimates

Most parameters either covered 0

\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[height=343px]{../DiseaseAnalysis/images/parplot_fmagna.pdf}
\caption{\textit{F. magna} model parameter estimates.}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[height=400px]{../DiseaseAnalysis/images/parplot_ptenuis.pdf}
\caption{\textit{P. tenuis} model parameter estimates.}
\end{subfigure}
\caption{\label{fig:parsplots}Displayed are the mode and 95\% credible interval for coefficients of fixed effects. See Appendix A for exact values.}
\end{figure}


## Bayesian model checks

We checked the fit of the models using Bayesian predictive checks made available
by the `INLA` package; specifically the conditional predictive ordinate (CPO)
and the probability integral transform (PIT).

### Conditional predictive ordinate

The conditional predictive ordinate (CPO) is a leave-one-out cross-validation measure,
defined as

$$
\text{CPO}_i = \pi(y_i^{\text{obs}} | y_{-i})
$$

and describes the probability of observing of a particular observation when all
of the data but observation $i$ are used to fit to the model. Very small values
indicate that the model is not likely to have predicted a particular observation
and thus that model is not expected to generalize well.

```{r cpo_get}

```

All of our models exhibited poor performance with roughly half of the data
points having low CPO values less than 0.1 (Figure \ref{fig:gofplot1}). In
addition, the data points with low CPO values are strongly related to those
points with the response > 0 (Figure \ref{fig:gofplot2}). This indicates that the models overall fit the
data poorly, having almost no predictive power to the domain of the response of
interest.

\begin{figure}[H]
\begin{subfigure}[b]{0.4\textwidth}
\centering
\includegraphics[height=400px]{../DiseaseAnalysis/images/cpo_fmagna.pdf}
\caption{\textit{F. magna} model goodness-of-fit statistics.}
\end{subfigure}
\begin{subfigure}[b]{0.6\textwidth}
\centering
\includegraphics[height=400px]{../DiseaseAnalysis/images/cpo_ptenuis.pdf}
\caption{\textit{P. tenuis} model goodness-of-fit statistics.}
\end{subfigure}
\caption{\label{fig:gofplot1}Displayed are the densities of two goodness-of-fit statistics for the models applied to the two parasite species' data. Under good model fit, the CPO values should not have many values approaching zero; the PIT values should be uniformly distributed.}
\end{figure}


### Probability integral transform

Another goodness-of-fit measure that `INLA` provides is the probability integral
transform value (PIT), which measures the probability of predicting a new
observation less than or equal to the observed data value. More generally the
probability integral transform states that the cumulative distribution function
$\mathcal{F}$ of a random variable $X$ evaluated at $x_i$ for all $i$ will have
a random uniform distribution if the random values $x_i$ are indeed generated by
the same process as $X$. If we take $\mathcal{F}(X)$ to be predictions from a
fitted model, departure from uniformity indicates that the observations $x_i$
may not be well predicted by the model.

\begin{figure}[H]
\begin{subfigure}[b]{0.9\textwidth}
\centering
\includegraphics[height=400px]{../DiseaseAnalysis/images/cpo_by_response_fmagna.pdf}
\caption{\textit{F. magna} model goodness-of-fit statistics.}
\end{subfigure}

\begin{subfigure}[b]{0.9\textwidth}
\centering
\includegraphics[height=400px]{../DiseaseAnalysis/images/cpo_by_response_ptenuis.pdf}
\caption{\textit{P. tenuis} model goodness-of-fit statistics.}
\end{subfigure}
\caption{\label{fig:gofplot2}Displayed are the conditional predictive ordinate and probbability transform values across values of the response variable. The goodness-of-fit values are relatively poor when the response is not zero.}
\end{figure}

By visual assessment, the PIT values are not uniformly distributed (Figure
\ref{fig:gofplot1}), and become more concentrated around 1 with increasing
values of the response (Figure \ref{fig:gofplot2}).

## Binary model results

Due to the poor fit, the models were reassessed with the data converted to
binary presence-absence data, and the best-performing models by WAIC were
re-assessed. The conditional predictive ordinate values were improved, having
few values below 0.1. However, the probability integral transform continued to
indicate a lack of fit, having a high density of values approaching unity.

\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[height=400px]{../DiseaseAnalysis/images/logreg_full_cpo.png}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[height=400px]{../DiseaseAnalysis/images/logreg_full_pit.png}
\caption{}
\end{subfigure}
\caption{\label{fig:gofplot2}Displayed are the (a) conditional predictive ordinate and (b) probability transform values.}
\end{figure}


# Discussion

# References

<div id=refs></div>

# Appendices

## Appendix A: Parameter estimates for the negative binomial model

```{r partable}
load('../DiseaseAnalysis/model_outputs/fmagna_effects.Rdata')
load('../DiseaseAnalysis/model_outputs/ptenuis_effects.Rdata')

fmfix = fmagna_effects$Fixed %>% select(variable, mean:mode)
ptfix = ptenuis_effects$Fixed %>% select(variable, mean:mode)

rle_fm = rle(fmagna_effects$Fixed$pubnames) %>% do.call(data.frame, args = .)
index_names = rle_fm$values %>% as.character
# index_lengths = Reduce(f = sum, x = rle_fm$lengths, accumulate = T)
index_lengths = rle_fm$lengths
index = index_lengths
names(index) = index_names
kable(x = fmfix, format = 'latex', digits = 3, booktabs = T, longtable = T, caption = "\\label{tab:fmtable} Fixed effect parameter estimates and distribution for \\textit{F. magna}") %>% 
  pack_rows(index = index) %>% 
  kable_styling(latex_options = c("repeat_header"))

```

```{r partable2}
rle_pt = rle(ptenuis_effects$Fixed$pubnames) %>% do.call(data.frame, args = .)
index_names = rle_pt$values %>% as.character
# index_lengths = Reduce(f = sum, x = rle_pt$lengths, accumulate = T)
index_lengths = rle_pt$lengths
index = index_lengths
names(index) = index_names
kable(x = ptfix, format = 'latex', digits = 3, booktabs = T, longtable = T, caption = "\\label{tab:pttable} Fixed effect parameter estimates and distribution for \\textit{P. tenuis}") %>% 
  pack_rows(index = index) %>% 
  kable_styling(latex_options = c("repeat_header"))

```

```{r raneftable}

fmagna_effects$Hyperpars %<>% left_join(model_matches, by = c("name" = "fitnames"))
ptenuis_effects$Hyperpars %<>% left_join(model_matches, by = c("name" = "fitnames"))
fmranef = fmagna_effects$Hyperpars  %>% select(variable, mean:mode)
fmranef$variable %<>% gsub(x = fmranef$variable, pattern = "size for the nbinomial observations \\(1\\/overdispersion\\)", replacement = "Nbinomial disp. par.", perl = T)

ptranef = ptenuis_effects$Hyperpars %>% select(variable, mean:mode) 
ptranef$variable %<>% gsub(x = ptranef$variable, pattern = "size for the nbinomial observations \\(1\\/overdispersion\\)", replacement = "Nbinomial disp. par.", perl = T)

rle_fm = rle(fmagna_effects$Hyperpars$pubnames)
index_lengths = rle_fm$lengths
index = index_lengths
names(index) = index_names

kable(x = fmranef, format = 'latex', digits = 3, booktabs = T, longtable = T, caption = "\\label{tab:fmtable2} Random effect parameter estimates and distribution for \\textit{F. magna}") %>% 
  pack_rows(index = index) %>% 
  kable_styling(latex_options = c("repeat_header")) %>%
  column_spec(1, width = "100px")
```

```{r raneftable2}
rle_pt = rle(ptenuis_effects$Hyperpars$pubnames)
index_lengths = rle_pt$lengths
index = index_lengths
names(index) = index_names

kable(x = ptranef, format = 'latex', digits = 3, booktabs = T, longtable = T, caption = "\\label{tab:pttable2} Random effect parameter estimates and distribution for \\textit{P. tenuis}") %>% 
  pack_rows(index = index) %>% 
  kable_styling(latex_options = c("repeat_header")) %>%
  column_spec(1, width = "100px")
```

